---
title: "An Introduction to CHNOSZ"
author: "Jeffrey M. Dick"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html:
    tufte_features: ["background"]
    toc: true
    mathjax: null
vignette: >
  %\VignetteIndexEntry{An Introduction to CHNOSZ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: vig.bib
link-citations: yes
csl: elementa.csl
---

```{css, echo=FALSE}
html { 
  font-size: 14px;
}
/* zero margin around pre blocks (looks more like R console output) */
pre {
  margin: 0;
  padding: 0;
}
</style>
```

```{r setup, include=FALSE}
library(knitr)

## From "Tufte Handout" example dated 2016-12-27
# Invalidate cache when the tufte version changes
opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)

## Setup pngquant to optimize PNG images
knit_hooks$set(pngquant = hook_pngquant)
pngquant <- "--speed=1 --quality=0-25"
# pngquant isn't available on R-Forge ...
if (!nzchar(Sys.which("pngquant"))) pngquant <- NULL 

## Colorize messages 20171031
## Adapted from https://gist.github.com/yihui/2629886#file-knitr-color-msg-rnw
color_block = function(color) {
  function(x, options) sprintf('<pre style="color:%s">%s</pre>', color, x)
}
knit_hooks$set(warning = color_block('magenta'), error = color_block('red'), message = color_block('blue'))
```

# CHNOSZ: An Introduction to Thermodynamic Calculations in R

This vignette introduces CHNOSZ, an R package for thermodynamic calculations relevant to geochemistry and geobiochemistry. CHNOSZ provides functions and a thermodynamic database for calculating properties of reactions involving minerals, aqueous species, and gases across a range of temperatures and pressures.

## Getting Started

After installing CHNOSZ from CRAN, load the package:

```{r library_CHNOSZ}
library(CHNOSZ)
```

This makes the thermodynamic database and functions available in your R session. To restore default settings at any point, use `reset()`.

## Core Functionality

CHNOSZ offers several primary functions for thermodynamic analysis:

### Functions Without Side Effects (Return Values)

* `info()`: Search the thermodynamic database
* `subcrt()`: Calculate thermodynamic properties of species and reactions
* `affinity()`: Calculate affinities of formation reactions
* `equilibrate()`: Calculate equilibrium chemical activities
* `diagram()`: Plot the results

### Functions With Side Effects (Modify System State)

* `basis()`: Set basis species and their chemical activities
* `species()`: Set species of interest and their activities
* `reset()`: Reset database and system settings to defaults

## Querying the Thermodynamic Database

The `info()` function provides access to the OBIGT thermodynamic database.

```{r info_CH4}
# Get database index for aqueous methane
info("CH4")
```

When searching by formula, aqueous species are returned if they are available.
Use a species name or add the state to get a particular physical state - `aq`, `cr`, `gas`, or `liq`:

```{r info_methane}
# Two ways to lookup methane gas
info("methane")
info("CH4", "gas")
```

Use `info()` recursively to return thermodynamic parameters:

```{r info_info_CH4}
# Get thermodynamic properties for aqueous methane
info(info("CH4"))
```

You can access fuzzy search functionality by using partial names:

```{r info_fuzzy}
# Search for ribose-related entries
info("ribose+")
```

## Calculating Thermodynamic Properties

The `subcrt()` function [named after SUPCRT; @JOH92] calculates standard thermodynamic properties:

```{r subcrt_CH4}
# Properties of aqueous methane at default T and P
subcrt("CH4")
```

<p></p>
```{r subcrt_TP}
# Custom T,P grid for water in supercritical region
subcrt("H2O", T = c(400, 500), P = c(250, 300))
```

Unit conversions are handled by `T.units()`, `P.units()`, and `E.units()`:

```{r E.units}
# Change energy units from Joules to calories
E.units("cal")
subcrt("CH4", T = 25)
reset()  # Restore defaults
```

## Working with Reactions

Define reactions with species names, states (optional), and coefficients:

```{r subcrt_CO2_reaction}
# CO2 dissolution reaction
subcrt(c("CO2", "CO2"), c("gas", "aq"), c(-1, 1), T = 25)
```

Reactions can be automatically balanced using basis species:

```{r basis}
basis(c("CO2", "H2O", "H+", "e-"))
species(c("CH4", "acetate"))
subcrt(c("CH4", "acetate"), c("aq", "aq"), c(1, -1), T = 25)
```

## Chemical Affinity and Stability Diagrams

```{r figure-setup, include = FALSE}
knitr::opts_chunk$set(
  fig.margin=TRUE, fig.width=6, fig.height=4, out.width="100%", results="hide",
  message=FALSE, cache=TRUE, pngquant=pngquant,
  # Set dpi 20231129
  dpi = if(nzchar(Sys.getenv("CHNOSZ_BUILD_LARGE_VIGNETTES"))) 72 else 50
)
```

The `affinity()` function calculates chemical affinities over ranges of T, P, and activities:

```{r diagram, fig.cap = "Eh-pH (Pourbaix) diagram for S"}
# Set up the C-H-N-O-S basis system with electron
basis("CHNOSe")
# Define aqueous sulfur species
species(c("SO4-2", "HSO4-", "HS-", "H2S"))
# Calculate affinities on an Eh-pH grid
a <- affinity(pH = c(0, 14), Eh = c(-0.8, 1.2))
# Create an Eh-pH (Pourbaix) diagram
diagram(a, fill = "terrain")
```

For more sophisticated diagrams involving speciation of basis species, use the `mosaic()` function:

```{r mosaic, fig.cap = "Cu-S-Cl-H2O mosaic diagram"}
# Create a mosaic diagram for Cu-S-Cl-H2O system
basis(c("Cu", "H2S", "Cl-", "H2O", "H+", "e-"))
basis("H2S", -6)
basis("Cl-", -1)
species(c("CuCl", "CuCl2-", "CuCl3-2", "CuCl+", "CuCl2", "CuCl3-", "CuCl4-2"))
species(c("chalcocite", "tenorite", "cuprite", "copper"), add = TRUE)
bases <- c("H2S", "HS-", "HSO4-", "SO4-2")
m <- mosaic(bases, pH = c(0, 12), Eh = c(-1, 1), T = 200)
diagram(m$A.species, lwd = 2, bold = species()$state == "cr")
diagram(m$A.bases, add = TRUE, col = 4, col.names = 4, italic = TRUE)
water.lines(m$A.species, lty = 3, col = 8)
```

## Equilibrium Calculations

Calculate equilibrium distributions of species:

```{r equilibrate, fig.cap="Carbonate speciation as a function of pH and temperature"}
# Carbonate speciation vs pH
basis(c("CO2", "H2O", "H+", "e-"))
species(c("CO2", "HCO3-", "CO3-2"))
# 25 degrees C
a <- affinity(pH = c(0, 14))
e <- equilibrate(a)
diagram(e, alpha = TRUE)
# 100 degrees C
a <- affinity(pH = c(4, 12), T = 100)
e <- equilibrate(a)
diagram(e, alpha = TRUE, add = TRUE, col = 2, names = NA)
```

Calculate solubility of minerals or gases:

```{r solubility, fig.cap="Solubility of corundum (green line) and equilibrium concentrations of aqueous species (black lines)"}
# Corundum solubility vs pH
basis(c("Al+3", "H2O", "H+", "e-"))
species("corundum")
iaq <- info(c("Al(OH)4-", "Al(OH)3", "Al(OH)2+", "AlOH+2", "Al+3"))
s <- solubility(iaq, pH = c(2, 10), in.terms.of = "Al+3")
diagram(s, col = 3, lwd = 2, ylim = c(-10, -2))
diagram(s, type = "loga.equil", add = TRUE)
legend("topright", c("25 째C", "1 bar"), bty = "n")
```

## Activity Coefficients

Incorporate non-ideal behavior using the extended Debye-H체ckel equation by setting the ionic strength parameter `IS`:

```{r solubility_IS, fig.cap="Solubility of corundum dependent on ionic strength"}
# Corundum solubility again
basis(c("Al+3", "H2O", "H+", "e-"))
species("corundum")
iaq <- info(c("Al(OH)4-", "Al(OH)3", "Al(OH)2+", "AlOH+2", "Al+3"))
# Calculate with ionic strength of 0 and 1 molal
s0 <- solubility(iaq, pH = c(2, 10))
s1 <- solubility(iaq, pH = c(2, 10), IS = 1)
diagram(s1, col = 4, lwd = 3, ylim = c(-8, -2))
diagram(s0, col = 3, lwd = 2, add = TRUE)
legend("topleft", legend = c(1, 0), lwd = c(3, 2), col = c(4, 3), title = "I (mol/kg)", bty = "n")
legend("topright", c("25 째C", "1 bar"), bty = "n")
```

Functions that have the `IS` parameter are `subcrt()`, `affinity()`, `mosaic()`, and `solubility()`.
When a value for `IS` is specified, inputs and outputs labeled as activities are actually interpreted or reported as molalities.

## Building Up

Having seen basic examples of the main features of CHNOSZ, here are ideas for building more complex calculations and diagrams.

### 1. Use retrieve() to search species by elements

Want to find all the Al complexes in the database?
List them by calling `retrieve()` with the main element optionally followed by the ligand elements and state.

```{r retrieve_Al_default, results = "show"}
# List aqueous Al species in the default database
iaq <- retrieve("Al", ligands = c("F", "H", "O"), state = "aq")
# Print the first few rows and columns
info(iaq)[1:3, 1:5]
# Use the species index or name in subcrt()
subcrt(iaq[1:3], T = 25)
#subcrt(names(iaq)[1:3], T = 25)  # same as above
```

The result of `retrieve()` can also be used to add species to a diagram; see the example in #3 below.

```{r Al_diagram}
basis(c("Al+3", "H2O", "F-", "H+", "e-"))
species(iaq)
species(names(iaq))  # same as above
```

### 2. Load optional data with add.OBIGT()

Perhaps you'd like to use data from older databases that have been superseded by later updates.
The OBIGT vignette briefly summarizes the superseded data for [SUPCRT92](OBIGT.html#optional-SUPCRT92) and [SLOP98](OBIGT.html#optional-SLOP98) [@SLOP98].
Use add.OBIGT() to load these old data entries.

```{r add_OBIGT_SLOP98, results = "show"}
add.OBIGT("SLOP98")
iaq_all <- retrieve("Al", ligands = c("F", "H", "O"), state = "aq")
# Use difference between sets to get the added species
info(setdiff(iaq_all, iaq))
```

The convention for SUPCRT-family databases is to use anhydrous species.
For example, AlO2- in SLOP98 corresponds to Al(OH)4- in the default database (see output above).
They are effectively the same species, which is why only the latter [taken from a more extensive compilation for Al species properties; @TS01] is used in the default database.
Unless you have a specific reason to compare them, redundant species should not be used in the same equilibrium calculation.

### 3. Use OBIGT() or reset() to restore the default database and settings

`OBIGT()` restores just the database, without affecting other settings (e.g. `E.units()`, `basis()` and `species()`).
`reset()` resets all settings in CHNOSZ, including the database.
These functions are useful for both interactive use and scripts that compare different versions of data or plots for different systems or conditions.

Let's put items #1-3 together to remake the corundum solubility plot using only species available in SLOP98.
To do this, we use `add.OBIGT()` followed by `retrieve()` to gather the species indices for all Al species, then taken only those species sourced from @SSWS97.

```{r corundum_solubility_2, fig.cap="Corundum solubility with species from SLOP98"}
# Add superseded species from SLOP98
add.OBIGT("SLOP98")
# List all aqueous Al species
iaq <- retrieve("Al", state = "aq")
# Keep only species from Shock et al. (1997)
iaq <- iaq[grepl("SSWS97", info(iaq)$ref1)]
# Plot corundum solubility vs pH
basis(c("Al+3", "H2O", "H+", "e-"))
species("corundum")
s <- solubility(iaq, pH = c(2, 10), in.terms.of = "Al+3")
## Alternatively, we could use the species names
#s <- solubility(names(iaq), pH = c(2, 10), in.terms.of = "Al+3")
diagram(s, col = 3, lwd = 2, ylim = c(-10, -2))
diagram(s, type = "loga.equil", add = TRUE)
legend("topright", c("25 째C", "1 bar"), bty = "n")
# Reset the database for subsequent calculations
reset()
```

4. Use `basis()` species to select compositional variables to plot

A common question is: what are the basis species used for?
The basis species define the compositional variables that can be added to a diagram.
In more precise terms, they define the thermodynamic components of a chemical system.
The composition of any possible species in that system can be represented by a linear combination of the basis species.

CHNOSZ requires that the number of basis species is equal to the number of different elements in the basis species (plus charge, if present).
If you were studying the relative stability of F- and OH-complexes with Al, you might be tempted to try this basis definition:

```{r basis_Al_F_OH_1, error=TRUE}
basis(c("Al+3", "F-", "OH-"))
```

According to the message, we don't have enough basis species for the number of elements.
Since hydroxide (OH-) is just water minus a proton, we could try this instead:

```{r basis_Al_F_OH_2, error=TRUE}
basis(c("Al+3", "F-", "H+", "H2O"))
```

That's still not enough species.
As is often the case, we need to include a basis species representing oxidation-reduction (redox) reactions, even if there are no redox reactions between the formed species.
Here are two possible basis definitions that do not give an error.

```{r basis_Al_F_OH_3}
# Use "oxygen" to get oxygen gas (for logfO2 diagrams)
basis(c("Al+3", "F-", "H+", "H2O", "oxygen"))
# Use "e-" to get aqueous electron (for Eh diagrams)
basis(c("Al+3", "F-", "H+", "H2O", "e-"))
```

5. Set activities of formed `species()` to define a single solubility contour

In order to make a diagram with stability fields for different species, CHNOSZ needs to know about the activities of all the species in the reaction.
The activities of the basis species start with constant values as shown in the output above (`logact` column).
Selected basis species can be assigned to plot axes (with a range of values) in `affinity()`.

How about the formed species in the system - that is, the species whose stability fields we want to visualize?
We both list the species and set their activities using `species()`.
The function defaults to activities of 1e-3 (`logact` of -3) for aqueous species and unit activity (`logact` = 0) for minerals, gases, and liquids.
Let's change this to activities of 1e-6 for the formed species.

```{r species_logact}
basis(c("Al+3", "F-", "H+", "H2O", "e-"))
iaq <- retrieve("Al", ligands = c("F", "H", "O"), state = "aq")
# Check that the data are from the same source
stopifnot(all(info(iaq)$ref1 == "TS01"))
species(iaq, -6)
```

This value for `logact` defines a solubility contour, as we'll see below.

6. When to use add = TRUE

There are two places where you might see `add = TRUE`.
First, in `species()` to add species not already in the list.
Without `add = TRUE`, any existing species are discarded.
Second, in `diagram()` to add to an existing plot.

Let's put items #4-6 together to make a Pourbaix (Eh-pH) diagram for Al with two solubility contours.

```{r Pourbaix_Mn, fig.cap = "Pourbaix diagram for Mn with two solubility contours"}
basis(c("Mn+2", "H+", "H2O", "e-"))
icr <- retrieve("Mn", ligands = c("H", "O"), state = "cr")
iaq <- retrieve("Mn", ligands = c("H", "O"), state = "aq")
# First layer, logact(aq) = -3
species(icr)
species(iaq, add = TRUE)
a <- affinity(pH = c(4, 14), Eh = c(-1, 1), T = 100)
# Use names = NA to avoid plotting labels twice
diagram(a, lty = 2, names = NA)
# Second layer, logact(aq) = -4
species(icr)
species(iaq, -4, add = TRUE)
a <- affinity(pH = c(4, 14), Eh = c(-1, 1), T = 100)
d <- diagram(a, add = TRUE)
# Add water stability limits
water.lines(d, lty = 3, col = 8)
# Add legends
legend("topright", legend = c(lT(100), lP("Psat")), bty = "n")
title = as.expression(quote(log~italic(a)))
legend("bottomleft", legend = c(-3, -4), lty = c(2, 1), title = title, bty = "n")
```

The shaded areas in the diagram represent water instability regions and are automatically added by `diagram()`.
We use `water.lines()` here to plot the water stability limits with dotted lines.

7. Set grid resolution and constant T, P, or IS in affinity()
8. Use solubility() to draw multiple solubility contours
9. Use the transect mode of affinity() for arbitrary numbers of variables
10. Choose non-default balancing constraints in diagram()
11. Extract lines and image data from the output of diagram()
12. Use helper functions to create labels for diagrams
13. Use the output of subcrt() to format reactions for diagrams
14. Calculate affinity values (opposite of Gibbs energy of reaction) with subcrt()
15. On-demand database updates with mod.OBIGT() and logK.to.OBIGT()

## Further Resources

- Browse the package documentation with `help(package = "CHNOSZ")`
- See function-specific help: `?info`, `?subcrt`, etc.
- Explore demos with `demo(package = "CHNOSZ")`

Additional vignettes cover:

- OBIGT thermodynamic database
- Creating multi-metal diagrams
- Customizing the thermodynamic database

## References

The CHNOSZ package incorporates data and methods from numerous sources. To view citation information:

```{r refs, eval = FALSE}
# View all references
thermo.refs()

# View references for specific species
thermo.refs(info("CH4"))
```

## Document history

- 2010-09-30 First version, titled "Getting started with CHNOSZ".
- 2017-02-15 Rewritten and switched from Sweave to [knitr](https://yihui.org/knitr/).
- 2025-04-07 Revised and shortened with AI assistance (except for [Building Up](#building-up)).

<p>
```{r the_end}
   ######    ##   ##    ##   ##    ######     #####  #####
 ##         ##===##    ## \\##   ##    ##     \\       //
 ######    ##   ##    ##   ##    ######    #####      #####
```
</p>

<!-- for finding what versions of packages are on R-Forge and winbuilder
```{r sessionInfo}
sessionInfo()
```
-->
