---
title: "An Introduction to CHNOSZ"
author: "Jeffrey M. Dick"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html:
    tufte_features: ["background"]
    toc: true
    mathjax: null
vignette: >
  %\VignetteIndexEntry{An Introduction to CHNOSZ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: vig.bib
csl: elementa.csl
---

```{css, echo=FALSE}
html { 
  font-size: 14px;
}
/* zero margin around pre blocks (looks more like R console output) */
pre {
  margin: 0;
  padding: 0;
}
</style>
```

```{r setup, include=FALSE}
library(knitr)

## From "Tufte Handout" example dated 2016-12-27
# Invalidate cache when the tufte version changes
opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)

## Setup pngquant to optimize PNG images
knit_hooks$set(pngquant = hook_pngquant)
pngquant <- "--speed=1 --quality=0-25"
# pngquant isn't available on R-Forge ...
if (!nzchar(Sys.which("pngquant"))) pngquant <- NULL 

## Colorize messages 20171031
## Adapted from https://gist.github.com/yihui/2629886#file-knitr-color-msg-rnw
color_block = function(color) {
  function(x, options) sprintf('<pre style="color:%s">%s</pre>', color, x)
}
knit_hooks$set(warning = color_block('magenta'), error = color_block('red'), message = color_block('blue'))
```

# CHNOSZ: An Introduction to Thermodynamic Calculations in R

This vignette introduces CHNOSZ, an R package for thermodynamic calculations relevant to geochemistry and geobiochemistry. CHNOSZ provides functions and a thermodynamic database for calculating properties of reactions involving minerals, aqueous species, and gases across a range of temperatures and pressures.

## Getting Started

After installing CHNOSZ from CRAN, load the package:

```{r library_CHNOSZ}
library(CHNOSZ)
```

This makes the thermodynamic database and functions available in your R session. To restore default settings at any point, use `reset()`.

## Core Functionality

CHNOSZ offers several primary functions for thermodynamic analysis:

### Functions Without Side Effects (Return Values)

* `info()`: Search the thermodynamic database
* `subcrt()`: Calculate thermodynamic properties of species and reactions
* `affinity()`: Calculate affinities of formation reactions
* `equilibrate()`: Calculate equilibrium chemical activities
* `diagram()`: Plot the results

### Functions With Side Effects (Modify System State)

* `basis()`: Set basis species and their chemical activities
* `species()`: Set species of interest and their activities
* `reset()`: Reset database and system settings to defaults

## Querying the Thermodynamic Database

The `info()` function provides access to the OBIGT thermodynamic database:

```{r info_CH4}
# Get database index for aqueous methane
info("CH4")
```

```{r info_info_CH4}
# Get thermodynamic properties for methane
info(info("CH4"))
```

You can access fuzzy search functionality by using partial names:

```{r info_fuzzy}
# Search for ribose-related entries
info("ribose+")
```

## Calculating Thermodynamic Properties

The `subcrt()` function (named after SUPCRT) calculates standard thermodynamic properties:

```{r subcrt_CH4}
# Properties of aqueous methane at default T and P
subcrt("CH4")
```

<p></p>
```{r subcrt_TP}
# Custom T,P grid for water in supercritical region
subcrt("H2O", T = c(400, 500), P = c(250, 300))
```

Unit conversions are handled by `T.units()`, `P.units()`, and `E.units()`:

```{r E.units}
# Convert energy units from Joules to calories
E.units("cal")
subcrt("CH4", T = 25)
reset()  # Restore defaults
```

## Working with Reactions

Define reactions with species names, states (optional), and coefficients:

```{r subcrt_CO2_reaction}
# CO2 dissolution reaction
subcrt(c("CO2", "CO2"), c("gas", "aq"), c(-1, 1), T = 25)
```

Reactions can be automatically balanced using basis species:

```{r basis}
basis(c("CO2", "H2O", "H+", "e-"))
species(c("CH4", "acetate"))
subcrt(c("CH4", "acetate"), c("aq", "aq"), c(1, -1), T = 25)
```

## Chemical Affinity and Stability Diagrams

```{r figure-setup, include = FALSE}
knitr::opts_chunk$set(
  fig.margin=TRUE, fig.width=6, fig.height=4, out.width="100%", results="hide",
  message=FALSE, cache=TRUE, pngquant=pngquant,
  # Set dpi 20231129
  dpi = if(nzchar(Sys.getenv("CHNOSZ_BUILD_LARGE_VIGNETTES"))) 72 else 50
)
```

The `affinity()` function calculates chemical affinities over ranges of T, P, and activities:

```{r diagram, fig.cap = "Eh-pH (Pourbaix) diagram for S"}
# Set up the C-H-N-O-S basis system with electron
basis("CHNOSe")
# Define aqueous sulfur species
species(c("SO4-2", "HSO4-", "HS-", "H2S"))
# Calculate affinities on an Eh-pH grid
a <- affinity(pH = c(0, 14), Eh = c(-0.8, 1.2))
# Create an Eh-pH (Pourbaix) diagram
diagram(a, fill = "terrain")
```

For more sophisticated diagrams involving speciation of basis species, use the `mosaic()` function:

```{r mosaic, fig.cap = "Cu-S-Cl-H2O mosaic diagram"}
# Create a mosaic diagram for Cu-S-Cl-H2O system
basis(c("Cu", "H2S", "Cl-", "H2O", "H+", "e-"))
basis("H2S", -6)
basis("Cl-", -1)
species(c("CuCl", "CuCl2-", "CuCl3-2", "CuCl+", "CuCl2", "CuCl3-", "CuCl4-2"))
species(c("chalcocite", "tenorite", "cuprite", "copper"), add = TRUE)
bases <- c("H2S", "HS-", "HSO4-", "SO4-2")
m <- mosaic(bases, pH = c(0, 12), Eh = c(-1, 1), T = 200)
diagram(m$A.species, lwd = 2, bold = species()$state == "cr")
diagram(m$A.bases, add = TRUE, col = 4, col.names = 4, italic = TRUE)
water.lines(m$A.species)
```

## Equilibrium Calculations

Calculate equilibrium distributions of species:

```{r equilibrate, fig.cap="Carbonate speciation as a function of pH and temperature"}
# Carbonate speciation vs pH
basis(c("CO2", "H2O", "H+", "e-"))
species(c("CO2", "HCO3-", "CO3-2"))
# 25 degrees C
a <- affinity(pH = c(0, 14))
e <- equilibrate(a)
diagram(e, alpha = TRUE)
# 100 degrees C
a <- affinity(pH = c(4, 12), T = 100)
e <- equilibrate(a)
diagram(e, alpha = TRUE, add = TRUE, col = 2, names = NA)
```

Calculate solubility of minerals or gases:

```{r solubility, fig.cap="Solubility of corundum (green line) and equilibrium concentrations of aqueous species (black lines)"}
# Corundum solubility vs pH
basis(c("Al+3", "H2O", "H+", "e-"))
species("corundum")
iaq <- info(c("Al(OH)4-", "Al(OH)3", "Al(OH)2+", "AlOH+2", "Al+3"))
s <- solubility(iaq, pH = c(2, 10), in.terms.of = "Al+3")
diagram(s, col = 3, lwd = 2, ylim = c(-10, -2))
diagram(s, type = "loga.equil", add = TRUE)
legend("topright", c("25 °C", "1 bar"), bty = "n")
```

## Activity Coefficients

Incorporate non-ideal behavior using the extended Debye-Hückel equation by setting the ionic strength parameter `IS`:

```{r solubility_IS, fig.cap="Solubility of corundum dependent on ionic strength"}
# Corundum solubility again
basis(c("Al+3", "H2O", "H+", "e-"))
species("corundum")
iaq <- info(c("Al(OH)4-", "Al(OH)3", "Al(OH)2+", "AlOH+2", "Al+3"))
# Calculate with ionic strength of 0 and 1 molal
s0 <- solubility(iaq, pH = c(2, 10))
s1 <- solubility(iaq, pH = c(2, 10), IS = 1)
diagram(s1, col = 4, lwd = 3, ylim = c(-8, -2))
diagram(s0, col = 3, lwd = 2, add = TRUE)
legend("topleft", legend = c(1, 0), lwd = c(3, 2), col = c(4, 3), title = "I (mol/kg)", bty = "n")
legend("topright", c("25 °C", "1 bar"), bty = "n")
```

Functions that have the `IS` parameter are `subcrt()`, `affinity()`, `mosaic()`, and `solubility()`.
When a value for `IS` is specified, inputs and outputs labeled as activities are actually interpreted or reported as molalities.

## Further Resources

- Browse the package documentation with `help(package = "CHNOSZ")`
- See function-specific help: `?info`, `?subcrt`, etc.
- Explore demos with `demo(package = "CHNOSZ")`

Additional vignettes cover:

- OBIGT thermodynamic database
- Creating multi-metal diagrams
- Customizing the thermodynamic database

## References

The CHNOSZ package incorporates data and methods from numerous sources. To view citation information:

```{r refs, eval = FALSE}
# View all references
thermo.refs()

# View references for specific species
thermo.refs(info("CH4"))
```

## Document history

- 2010-09-30 Initial version, titled "Getting started with CHNOSZ".
- 2012-06-16 Add "More activity diagrams" (section no longer exists).
- 2015-05-14 Add warning about [internal consistency of thermodynamic data](#thermodynamic-database).
- 2017-02-15 Completely rewritten; switch from Sweave to [knitr](https://yihui.org/knitr/) ([Tufte style](https://rstudio.github.io/tufte/)).
- 2019-01-24 Add [section on solubility calculations](#complete-equilibrium-solubility).
- 2025-04-07 Shortened and revised with AI assistance.

<p>
```{r the_end}
   ######    ##   ##    ##   ##    ######     #####  #####
 ##         ##===##    ## \\##   ##    ##     \\       //
 ######    ##   ##    ##   ##    ######    #####      #####
```
</p>

<!-- for finding what versions of packages are on R-Forge and winbuilder
```{r sessionInfo}
sessionInfo()
```
-->
