---
title: "CHNOSZ FAQ"
author: "Jeffrey M. Dick"
output:
  html_vignette:
    mathjax: null
    toc: true
vignette: >
  %\VignetteIndexEntry{CHNOSZ FAQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: vig.bib
csl: elementa.csl
---

<style>
/* CSS snippet from boostrap.css modified by jmd on 2023-06-23 */
/*!
 * Bootstrap  v5.3.0 (https://getbootstrap.com/)
 * Copyright 2011-2023 The Bootstrap Authors
 * Licensed under MIT (https://github.com/twbs/bootstrap/blob/main/LICENSE)
 */
:root,
[data-bs-theme=light] {
  --bs-info-text-emphasis: #31708f;
  --bs-info-bg-subtle: #d9edf7;
  --bs-info-border-subtle: #9eeaf9;
  --bs-border-width: 1px;
  --bs-border-radius: 0.375rem;
}
.alert {
  --bs-alert-bg: transparent;
  --bs-alert-padding-x: 0.5rem;
  --bs-alert-padding-y: 0.5rem;
  --bs-alert-margin-bottom: 1rem;
  --bs-alert-color: inherit;
  --bs-alert-border-color: transparent;
  --bs-alert-border: var(--bs-border-width) solid var(--bs-alert-border-color);
  --bs-alert-border-radius: var(--bs-border-radius);
  --bs-alert-link-color: inherit;
  position: relative;
  padding: var(--bs-alert-padding-y) var(--bs-alert-padding-x);
  margin-bottom: var(--bs-alert-margin-bottom);
  color: var(--bs-alert-color);
  background-color: var(--bs-alert-bg);
  border: var(--bs-alert-border);
  border-radius: var(--bs-alert-border-radius);
}
.alert-info {
  --bs-alert-color: var(--bs-info-text-emphasis);
  --bs-alert-bg: var(--bs-info-bg-subtle);
  --bs-alert-border-color: var(--bs-info-border-subtle);
  --bs-alert-link-color: var(--bs-info-text-emphasis);
}
</style>

<script>
function ToggleDiv(ID) {
  var D = document.getElementById("D-" + ID);
  var B = document.getElementById("B-" + ID);
  if (D.style.display === "none") {
    // open the div and change button text
    D.style.display = "block";
    B.innerText = "Hide code";
  } else {
    // close the div and change button text
    D.style.display = "none";
    B.innerText = "Show code";
  }
}
</script>

```{r setup, include = FALSE}
library(CHNOSZ)
options(width = 80)
# Use pngquant to optimize PNG images
library(knitr)
knit_hooks$set(pngquant = hook_pngquant)
pngquant <- "--speed=1 --quality=0-25"
if (!nzchar(Sys.which("pngquant"))) pngquant <- NULL
# To make warnings appear in text box 20230619
# https://selbydavid.com/2017/06/18/rmarkdown-alerts/
knitr::knit_hooks$set(
   error = function(x, options) {
     paste('\n\n<div class="alert alert-danger">',
           gsub('##', '\n', gsub('^##\ Error', '**Error:**', x)),
           '</div>', sep = '\n')
   },
   warning = function(x, options) {
     paste('\n\n<div class="alert alert-warning">',
           gsub('##', '\n', gsub('^##\ Warning:', '**Warning:**', x)),
           '</div>', sep = '\n')
   },
   message = function(x, options) {
     paste('\n\n<div class="alert alert-info">',
           gsub('##', '\n', x),
           '</div>', sep = '\n')
   }
)
```

```{r HTML, include = FALSE}
NOTE <- '<span style="background-color: yellow;">NOTE</span>'
# CHNOSZ functions
equilibrate_ <- '<code>equilibrate()</code>'
info_ <- '<code>info()</code>'
thermo.refs_ <- '<code>thermo.refs()</code>'
# Math stuff
logK <- "log&thinsp;<i>K</i>"
Hplus <- "H<sup>+</sup>"
HCO2_ <- "HCO<sub>2</sub><sup>−</sup>"
HCO3_ <- "HCO<sub>3</sub><sup>−</sup>"
O2 <- "O<sub>2</sub>"
log <- "log&thinsp;"
aHCO2_ <- "<i>a</i><sub>HCO<sub>2</sub><sup>−</sup></sub>"
aHCO3_ <- "<i>a</i><sub>HCO<sub>3</sub><sup>−</sup></sub>"
logfO2 <- "log&thinsp;<i>f</i><sub>O<sub>2</sub></sub>"
Ctot <- "C<sub>tot</sub>"
C3H5O2_ <- "C<sub>3</sub>H<sub>5</sub>O<sub>2</sub><sup>−</sup>"
a3HCO3_ <- "<i>a</i><sup>3</sup><sub>HCO<sub>3</sub><sup>−</sup></sub>"
aC3H5O2_ <- "<i>a</i><sub>C<sub>3</sub>H<sub>5</sub>O<sub>2</sub><sup>−</sup></sub>"
a2HCO3_ <- "<i>a</i><sup>2</sup><sub>HCO<sub>3</sub><sup>−</sup></sub>"
logCtot <- "log&thinsp;C<sub>tot</sub>"
CO2 <- "CO<sub>2</sub>"
H2O <- "H<sub>2</sub>O"
```

This vignette was compiled on `r Sys.Date()` with CHNOSZ version `r sessionInfo()$otherPkgs$CHNOSZ$Version`.

## How is 'CHNOSZ' pronounced?

As one syllable that starts with an *sh* sound and [rhymes with *Oz*](https://en.wiktionary.org/wiki/Rhymes:English/%C9%92z).
CHNOSZ and [schnoz](https://en.wiktionary.org/wiki/schnozz) are homophones.

*Answered on 2023-05-22.*

## How can CHNOSZ be cited?

* This paper is the general reference for CHNOSZ: @Dic19.
* This paper describes the features for making diagrams with multiple metals: @Dic21b.
* This paper describes the features for calculations involving proteins: @Dic08.
* The [<span style="color:blue">*OBIGT thermodynamic database*</span>](OBIGT.html) represents the work of many researchers.
**If you publish results that depend on any of these data, please cite the primary sources.**
Use `r info_` to show the reference keys for particular species and `r thermo.refs_` to list the bibliographic details:
```{r alanine, message = FALSE}
info(info("alanine"))[c("ref1", "ref2")]
thermo.refs(info("alanine"))
```

*Answered on 2023-05-27.*

## What thermodynamic models are used in CHNOSZ?

* The thermodynamic properties of liquid water are calculated using Fortran code from SUPCRT92 [@JOH92] or an implementation in R of the IAPWS-95 formulation [@WP02].
* Thermodynamic properties of other species are taken from a database for minerals and inorganic and organic aqueous species including biomolecules, or from amino acid group additivity for proteins [@DLH06].
* The corresponding high-temperature properties are calculated using the @BB85 equations for minerals and the revised Helgeson-Kirkham-Flowers [@HKF81] equations for aqueous species.
* The HKF equations are augmented with the Deep Earth Water (DEW) model [@SHA14] and estimates of parameters in the extended Debye-H&uuml;ckel equation [@MSS13] to calculate standard-state properties and activity coefficients for given ionic strength at high pressure (to 6 GPa).
* Activity coefficients are implemented via adjusted standard Gibbs energies at specified ionic strength [@Alb96], which converts all activity variables in the workflow to molalities.
* A related adjustment is available to convert standard Gibbs energies for gases from the 1 bar standard state used in SUPCRT to a variable-pressure standard state [@AC93,Ch.12].

*Answered on 2018-11-13; moved from <https://chnosz.net/> on 2023-05-27.*

## When and why do equal-activity boundaries depend on total activity?

Short answer: When the species have the same number of the conserved element (let's take C for example),
their activities are raised to the same exponent in reaction quotient, so the activity ratio in the law of mass action becomes unity.
But when the species have different numbers of the conserved element (for example, propanoate with 3 C and bicarbonate with 1 C),
their activities are raised to different exponents, and the activity ratio does not become unity even when the activities are equal
(except for the specific case where the activities themselves are equal to 1).
Therefore, in general, the condition of "equal activity" is not sufficient to define boundaries on a relative stability diagram;
instead, we need to say "activity of each species equal to *x*" or alternatively "total activity equal to *y*".

Long answer: First, consider a reaction between formate and bicarbonate: `r HCO2_` + 0.5 `r O2` $\rightleftharpoons$ `r HCO3_`.
The law of mass action (LMA) is <strong>`r logK` $=$ `r log`(`r aHCO3_` / `r aHCO2_`) $-$ 0.5 `r logfO2`</strong>.
The condition of equal activity is <font color="red">`r aHCO2_` $=$ `r aHCO3_`</font>.
Then, the LMA simplifies to <strong>`r logK` $=$ $-$ 0.5 `r logfO2`</strong>.
The total activity of C is given by <font color="blue">`r Ctot` $=$ `r aHCO2_` $+$ `r aHCO3_`</font>.
According to the LMA, `r logfO2` is a function only of `r logK`, so <font color="green">*d*`r logfO2`/*d*`r logCtot` $=$ 0</font>.
In other words, the position of the equal-activity boundary is independent of the value of `r Ctot`.

Next, consider a reaction between propanoate and bicarbonate: `r C3H5O2_` + <sup>7</sup>&frasl;<sub>2</sub> `r O2` $\rightleftharpoons$ 3 `r HCO3_` + 2 `r Hplus`.
The LMA is <strong>`r logK` $=$ `r log`(`r a3HCO3_` / `r aC3H5O2_`) $-$ pH $-$ <sup>7</sup>&frasl;<sub>2</sub> `r logfO2`</strong>.
The condition of equal activity is <font color="red">`r aC3H5O2_` $=$ `r aHCO3_`</font>.
Then, the LMA simplifies to <strong>`r logK` $=$ `r log``r a2HCO3_` $-$ pH $-$ <sup>7</sup>&frasl;<sub>2</sub> `r logfO2`</strong>.
The total activity of C is given by <font color="blue">`r Ctot` $=$ 3 `r aC3H5O2_` $+$ `r aHCO3_`</font>; combined with the condition of equal activity,
this gives <font color="blue">`r Ctot` $=$ 4 `r aHCO3_`</font>.
Substituting this into the LMA gives <strong>`r logK` $=$ `r log`(`r Ctot` / 4)<sup>2</sup> $-$ pH $-$ <sup>7</sup>&frasl;<sub>2</sub> `r logfO2`</strong>,
which can be rearranged to write `r logfO2` $=$ <sup>2</sup>&frasl;<sub>7</sub> (2 `r log``r Ctot` $-$ `r logK` $-$ `r log`16 $-$ pH).
It follows that <font color="green">*d*`r logfO2`/*d*`r logCtot` $=$ <sup>4</sup>&frasl;<sub>7</sub></font>,
and the position of the equal-activity boundary depends on `r Ctot`.

According to this analysis, increasing `r Ctot` from 0.03 to 3 molal (a 2 log-unit increase)
would have no effect on the location of the formate-bicarbonate equal-activity boundary,
but would raise the propanoate-bicarbonate equal-activity boundary by <sup>8</sup>&frasl;<sub>7</sub> units on the `r logfO2` scale.
Because the reaction between bicarbonate and `r CO2` does not involve `r O2` (but rather `r H2O` and `r Hplus`),
the same effect should occur on the propanoate-`r CO2` equal-activity boundary.
The plots below, which are made using `r equilibrate_` for species in the Deep Earth Water (DEW) model, illustrate this effect.

<button id="B-DEW_Ctot" onclick="ToggleDiv('DEW_Ctot')">Show code</button>
<div id="D-DEW_Ctot" style="display: none">
```{r DEW_Ctot, eval = FALSE}
### Based on demo/DEW.R in CHNOSZ

# Set up subplots
par(mfrow = c(1, 2), mar = c(3.0, 3.5, 2.5, 1.0), mgp = c(1.7, 0.3, 0), las = 1, tcl = 0.3, xaxs = "i", yaxs = "i")

# Activate DEW model
oldwat <- water("DEW")

###########
## logfO2-pH diagram for aqueous inorganic and organic carbon species at high pressure
## After Figure 1b of Sverjensky et al., 2014b [SSH14]
## (Nature Geoscience, https://doi.org/10.1038/NGEO2291)
###########

# Define system
basis("CHNOS+")
inorganic <- c("CO2", "HCO3-", "CO3-2", "CH4")
organic <- c("formic acid", "formate", "acetic acid", "acetate", "propanoic acid", "propanoate")
species(c(inorganic, organic), 0)
fill <- c(rep("aliceblue", length(inorganic)), rep("aquamarine", length(organic)))

# A function to make the diagrams
dfun <- function(T = 600, P = 50000, res = 200, Ctot = 0.03) {
  a <- affinity(pH = c(0, 10, res), O2 = c(-24, -12, res), T = T, P = P)
  # Set total C concentration to 0.03 molal
  # (from EQ3NR model for eclogite [Supporting Information of SSH14])
  e <- equilibrate(a, loga.balance = log10(Ctot))
  diagram(e, fill = fill)
  dp <- describe.property(c("     T", "     P"), c(T, P), digits = 0)
  legend("bottomleft", legend = dp, bty = "n")
}

water("DEW")
add.OBIGT("DEW")
dfun(Ctot = 0.03)
mtitle(c("Inorganic and organic species", "C[total] = 0.03 molal"))
dfun(Ctot = 3)
mtitle(c("Inorganic and organic species", "C[total] = 3 molal"))
```
</div>

```{r DEW_Ctot, echo = FALSE, message = FALSE, results = "hide", fig.width = 8, fig.height = 4, out.width = "100%", fig.align = "center", pngquant = pngquant}
```

*Answered on 2023-05-17.*

## How can minerals with phase transitions be added to the database?

Many minerals undergo phase transitions upon heating.
The stable phases at different temperatures are called polymorphs.
Each polymorph for a given mineral should have its own entry in OBIGT, including values of the standard thermodynamic properties (Δ*G*°, Δ*H*°, and *S*°) at 25 °C.
The 25 °C (or 298.15 K) values for high-temperature polymorphs are often not listed in thermodynamic tables, but they can be calculated.
This thermodynamic cycle shows how: we calculate the changes of a thermodyamic property (pictured here as `DS1` and `DS2`) between 298.15 K and the transition temperature (`Ttr`) for two polymorphs, then combine those with the property of the polymorphic transition (`DStr`) to obtain the difference of the property between the polymorphs at 298.15 K (`DS298`).

```text
               DStr              DStr: entropy of transition between polymorphs 1 and 2
      Ttr  o---------->o          Ttr: temperature of transition
           ^           |         
           |           |         
       DS1 |           | DS2      DS1: entropy change of polymorph 1 from 298.15 K to Ttr
           |           |          DS2: entropy change of polymorph 2 from 298.15 K to Ttr
           |           v
 298.15 K  o==========>o    DS298: entropy difference between polymorphs 1 and 2 at 298.15 K
               DS298        DS298 = DS1 + DStr - DS2
Polymorph  1           2
```

As an example, let's add pyrrhotite (Fe0.877S) from @PMW87.
The formula and thermodynamic properties of this pyrrhotite differ from those of FeS from @HDNB78, which is already in OBIGT.
We begin by defining all the input values in the next code block.
In addition to `G`, `H`, `S`, and the heat capacity coefficients, non-NA values of volume (`V`) must be provided for the polymorph transitions to be calculated correctly by `subcrt()`.

```{r pyrrhotite_values, message = FALSE}
# The formula of the new mineral
formula <- "Fe0.877S"
# Because the properties from Pankratz et al. (1987) are listed in calories,
# we need to change the output of subcrt() to calories (the default is Joules)
E.units("cal")
# Use temperature in Kelvin for the calculations below
T.units("K")
# Thermodynamic properties of polymorph 1 at 25 °C (298.15 K)
G1 <- -25543
H1 <- -25200
S1 <- 14.531
Cp1 <- 11.922
# Heat capacity coefficients for polymorph 1
a1 <- 7.510
b1 <- 0.014798
# For volume, use the value from Helgeson et al. (1978)
V1 <- V2 <- 18.2
# Transition temperature
Ttr <- 598
# Transition enthalpy (cal/mol)
DHtr <- 95
# Heat capacity coefficients for polymorph 2
a2 <- -1.709
b2 <- 0.011746
c2 <- 3073400
# Maximum temperature of polymorph 2
T2 <- 1800
```

Use the temperature (`Ttr`) and enthalpy of transition (`DHtr`) to calculate the entropy of transition (`DStr`).
Note that the Gibbs energy of transition (`DGtr`) is zero at `Ttr`.
```{r pyrrhotite_Ttr, message = FALSE}
DGtr <- 0  # DON'T CHANGE THIS
TDStr <- DHtr - DGtr  # TΔS° = ΔH° - ΔG°
DStr <- TDStr / Ttr
```

Put the heat capacity coefficients and volume in the database.
There is no tabulated value of `Cp` of the high-temperature polymorph extrapolated to 298.15 K, so leave it out.
If the properties were in Joules, we would omit `E_units = "cal"` or change it to `E_units = "J"`.
```{r pyrrhotite_Cp, results = "hide", message = FALSE}
mod.OBIGT("pyrrhotite_new", formula = formula, state = "cr", E_units = "cal",
  G = 0, H = 0, S = 0, V = V1, Cp = Cp1,
  a = a1, b = b1, c = 0, d = 0, e = 0, f = 0, lambda = 0, T = Ttr)
mod.OBIGT("pyrrhotite_new", formula = formula, state = "cr2", E_units = "cal",
  G = 0, H = 0, S = 0, V = V2,
  a = a2, b = b2, c = c2, d = 0, e = 0, f = 0, lambda = 0, T = T2)
```

For the time being, we set `G`, `H`, and `S` (i.e., the properties at 25 °C) to zero in order to easily calculate the temperature integrals of the properties from 298.15 K to `Ttr`.
Values of zero are placeholders that don't satisfy Δ*G*° = Δ*H*° − *T*Δ*S*° for the formation properties from the elements for either polymorph, as indicated by the following messages.
We will check again for self-consistency at the end of the example.
```{r pyrrhotite_info, results = "hide", collapse = TRUE}
info(info("pyrrhotite_new", "cr"))
info(info("pyrrhotite_new", "cr2"))
```

In order to calculate the temperature integrals of Δ*G*° and Δ*H*°, we need not only the heat capacity coefficients but also actual values of *S*°.
Therefore, we start by calculating the entropy changes of each polymorph from 298.15 to `r Ttr` K (`DS1` and `DS2`) and combining those with the entropy of transition to obtain the difference of entropy between the polymorphs at 298.15 K.
For polymorph 1 (with `state = "cr"`) it's advisable to include `use.polymorphs = FALSE` to prevent `subcrt()` from trying to identify the most stable polymorph at the indicated temperature.
```{r pyrrhotite_S, message = FALSE}
DS1 <- subcrt("pyrrhotite_new", "cr", P = 1, T = Ttr, use.polymorphs = FALSE)$out[[1]]$S
DS2 <- subcrt("pyrrhotite_new", "cr2", P = 1, T = Ttr)$out[[1]]$S
DS25 <- DS1 + DStr - DS2
```

Put the values of *S*° at 298.15 in OBIGT, then calculate the changes of all thermodynamic properties of each polymorph between 298.15 K and `Ttr`.
```{r pyrrhotite_D1_D2, message = FALSE, results = "hide"}
mod.OBIGT("pyrrhotite_new", state = "cr", S = S1)
mod.OBIGT("pyrrhotite_new", state = "cr2", S = S1 + DS25)
D1 <- subcrt("pyrrhotite_new", "cr", P = 1, T = Ttr, use.polymorphs = FALSE)$out[[1]]
D2 <- subcrt("pyrrhotite_new", "cr2", P = 1, T = Ttr)$out[[1]]
```

It's a good idea to check that the entropy of transition is calculated correctly.
```{r pyrrhotite_check_S, results = "hide"}
stopifnot(all.equal(D2$S - D1$S, DStr))
```

Now we're ready to add up the contributions to Δ*G*° and Δ*H*° from the left, top, and right sides of the cycle.
This gives us the differences between the polymorphs at 298.15 K, which we use to make the final changes to the database.
```{r pyrrhotite_DG25_DH25, results = "hide", message = FALSE}
DG25 <- D1$G + DGtr - D2$G
DH25 <- D1$H + DHtr - D2$H
mod.OBIGT("pyrrhotite_new", state = "cr", G = G1, H = H1)
mod.OBIGT("pyrrhotite_new", state = "cr2", G = G1 + DG25, H = H1 + DH25)
```

It's a good idea to check that the values of `G`, `H`, and `S` at 25 °C for each polymorph are self-consistent.
Here we use `check.GHS()` to calculate the difference between the value given for `G` and the value calculated from `H` and `S`.
The difference of less than 1 `r E.units()`/mol can probably be attributed to small differences in the entropies of the elements used by @PMW87 and in CHNOSZ.
We still get a message that the database value of `Cp` at 25 °C for the high-temperature polymorph is NA; this is OK because the (extrapolated) value can be calculated from the heat capacity coefficients.
```{r pyrrhotite_info2, collapse = TRUE}
cr_parameters <- info(info("pyrrhotite_new", "cr"))
stopifnot(abs(check.GHS(cr_parameters)) < 1)
cr2_parameters <- info(info("pyrrhotite_new", "cr2"))
stopifnot(abs(check.GHS(cr2_parameters)) < 1)
```

For the curious, here are the parameter values:
```{r pyrrhotite_parameters}
cr_parameters
cr2_parameters
```

Finally, let's look at the thermodynamic properties of the newly added pyrrhotite as a function of temperature around `Ttr`.
Here, we use the feature of `subcrt()` that identifies the stable polymorph at each temperature.
Note that Δ*G*° is a continuous function -- visual confirmation that the parameters yield zero for `DGtr` -- but Δ*H*°, *S*°, and *Cp*° are discontinuous at the transition temperature.

<button id="B-pyrrhotite_T" onclick="ToggleDiv('pyrrhotite_T')">Show code</button>
<div id="D-pyrrhotite_T" style="display: none">
```{r pyrrhotite_T, eval = FALSE}
T <- seq(550, 650, 1)
sout <- subcrt("pyrrhotite_new", T = T, P = 1)$out[[1]]
par(mfrow = c(1, 4), mar = c(4, 4.2, 1, 1))
labels <- c(G = "DG0", H = "DH0", S = "S0", Cp = "Cp0")
for(property in c("G", "H", "S", "Cp")) {
  plot(sout$T, sout[, property], col = sout$polymorph,
    xlab = axis.label("T"), ylab = axis.label(labels[property])
  )
  abline(v = Ttr, lty = 3, col = 8)
  if(property == "G")
    legend("bottomleft", c("1", "2"), pch = 1, col = c(1, 2), title = "Polymorph")
}
```
</div>
```{r pyrrhotite_T, echo = FALSE, message = FALSE, results = "hide", fig.width = 8, fig.height = 2.5, out.width = "100%", fig.align = "center", pngquant = pngquant}
```

For additional polymorphs, we could repeat the above procedure using polymorph 2 as the starting point to calculate `G`, `H`, and `S` of polymorph 3, and so on.

*Answered on 2023-06-23.*

## References
