---
title: "Diagrams with multiple metals"
author: "Jeffrey M. Dick"
output:
  html_vignette:
    mathjax: null
vignette: >
  %\VignetteIndexEntry{Diagrams with multiple metals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: vig.bib
csl: elementa.csl
---
<style>
/* https://gomakethings.com/how-to-break-an-image-out-of-its-parent-container-with-css/ */
@media (min-width: 700px) {
  .full-width {
    left: 50%;
    margin-left: -50vw;
    margin-right: -50vw;
    max-width: 100vw;
    position: relative;
    right: 50%;
    width: 100vw;
  }
}
@media (min-width: 1020px) {
  .full-width {
    left: 50vw; /* fallback if needed */
    left: calc(50vw - 160px);
    width: 1000px;
    position: relative;
    background-color: #9ecff7;
    padding:10px;
  }
}
/* zero margin around pre blocks (looks more like R console output) */
pre {
  margin-top: 0;
  margin-bottom: 0;
}
</style>
<script>
function ToggleDiv(ID) {
  var D = document.getElementById("D-" + ID);
  var B = document.getElementById("B-" + ID);
  if (D.style.display === "none") {
    // open the div and change button text
    D.style.display = "block";
    B.innerText = "Hide code";
  } else {
    // close the div and change button text
    D.style.display = "none";
    B.innerText = "Show code";
  }
}
</script>

```{r setup, include=FALSE}
## use pngquant to optimize PNG images
library(knitr)
knit_hooks$set(pngquant = hook_pngquant)
pngquant <- "--speed=1 --quality=0-25"
if (!nzchar(Sys.which("pngquant"))) pngquant <- NULL
```

```{r CHNOSZ_reset, include=FALSE}
library(CHNOSZ)
reset()
```

This vignette was compiled on `r Sys.Date()` with CHNOSZ version `r sessionInfo()$otherPkgs$CHNOSZ$Version`.

Basic diagrams in CHNOSZ are made for reactions that are *balanced on an element* (see [Equilibrium in CHNOSZ](equilibrium.html)) and therefore represent minerals or aqueous species that all have one element, often a metal, in common.
This vignette describes some methods for constructing diagrams for elements with multiple metals.
The methods are **simple overlay**, **mosaic series**, and **secondary balancing**.

## Simple Overlay

Simple overlay refers to independent calculations for two different systems that are displayed on the same diagram.
It is easy to make such a diagram, but there is no interaction between the systems.

## Mosaic Series

A mosaic diagram shows the effects of changing basis species on the stabilities of minerals.
The Fe-S-O-H system is a common example: the speciation of aqueous sulfur species affects the stabilities of iron oxides and sulfides.
Examples of mosaic diagrams with Fe or other single metals are given elsewhere.

A mosaic series is when predominance fields for minerals calculated in one mosaic diagram are used as input to a second mosaic diagram, where the minerals are now themselves basis species.
The example here shows the construction of a Cu-Fe-S-O-H diagram.

First we define the conditions and basis species.
It is important to put Cu^+^ first so that it will be used as the balance for the reactions with Cu-bearing minerals (which also have Fe).
Pyrite is chosen as the starting Fe-bearing basis species, which will be changed as indicated in `bases2`.

```{r series1, results = "hide", message = FALSE}
logaH2S <- -2
T <- 200
pH <- c(0, 12, 500)
O2 <- c(-50, -30, 500)
basis(c("Cu+", "pyrite", "H2S", "oxygen", "H2O", "H+"))
basis("H2S", logaH2S)
bases1 <- c("H2S", "HS-", "HSO4-", "SO4-2")
bases2 <- c("pyrite", "pyrrhotite", "magnetite", "hematite")
```

Now we calculate affinities for minerals in the Fe-S-O-H system using changing aqueous sulfur species in `bases1`.
The result is used to make different layers of the diagram:

1. Water stability region (bounded by the grey area)
2. Predominance fields for the aqueous S species (italic blue text and dashed lines)
3. Stability areas for the Fe-bearing minerals (black text and solid lines)

```{r series2, eval = FALSE, echo = 1:6}
species(bases2)
mFe <- mosaic(bases1, pH = pH, O2 = O2, T = T)
diagram(mFe$A.bases, lty = 0, names = NA)
diagram(mFe$A.bases, lty = 2, col = 4, col.names = 4,
        italic = TRUE, add = TRUE, limit.water = FALSE)
dFe <- diagram(mFe$A.species, add = TRUE, limit.water = FALSE)
species(c("chalcopyrite", "bornite"))
mCu <- mosaic(list(bases1, bases2), pH = pH, O2 = O2,
              T = T, predominant = list(NULL, dFe$predominant))
diagram(mCu$A.species, add = TRUE, col = 2, col.names = 2,
        lwd = 2, bold = TRUE, limit.water = FALSE)
TP <- describe.property(c("T", "P"), c(T, "Psat"))
legend("topright", TP, bty = "n")
title(paste("Cu-Fe-S-O-H; Total S =", 10^logaH2S, "m"))
```

Next we load the Cu-bearing minerals and calculate their affinities while changing *both* the aqueous sulfur species and the Fe-bearing minerals whose stability fields were just calculated.
The result is used to plot the final layer of the diagram:

4. Stability areas for Cu-bearing minerals (bold red text and solid lines)

After that we add the legend and title.

```{r series2, echo=7:14, results = "hide", message = FALSE, fig.width = 6, fig.height = 5, out.width = "80%", fig.align = "center", pngquant = pngquant}
```

This diagram has a distinctive chalcopyrite "hook" that is controlled on the low-pH side by the reaction with pyrite.
Only that reaction is shown in many published diagrams [e.g. @And75;@Gio02], but diagrams with a similar chalcopyrite wedge or hook can be seen in @BBR77 and @Bri80.

## Secondary Balancing

Predominance diagrams in CHNOSZ are made using the maximum affinity method, where the affinities of formation reactions of species are divided by the *balancing coefficients* [@Dic19].
Usually, these balancing coefficients are taken from the formation reactions themselves; for example, if they are the coefficients on the Fe-bearing basis species, then the reactions are said to be "balanced on Fe".

Some diagrams in the literature are made with secondary balancing constraints in addition to the primary ones.
For example, reactions of Fe-bearing minerals are balanced on Fe, and reactions of Cu-bearing minerals are balanced on Cu; these are both primary balancing coefficients.
Then, reactions between all minerals are balanced on H^+^ as the secondary balancing coefficients.
The core concept is to apply the secondary balance while also maintaining the primary balance; a method to do this has been implemented in `duplex()` function.

The example here makes a diagram like one shown in Figure 5 of @Bri80 and Figure 5 of @MH85.
Different parts of the script to make the diagrams are described below; press the button to show the entire script.

<button id="B-duplex" onclick="ToggleDiv('duplex')">Show code</button>
<div id="D-duplex" style="display: none">
```{r duplex, eval = FALSE}
mat <- matrix(c(1, 1, 2, 2, 3, 3, 4, 4, 4, 5, 5, 5), byrow = TRUE, nrow = 2)
layout(mat)
par(font.main = 1)

basis(c("Fe+2", "Cu+", "hydrogen sulfide", "oxygen", "H2O", "H+"))
xlab <- ratlab("Fe+2", "Cu+")

### PRIMARY balancing

# Only Fe-bearing minerals
species(c("pyrite", "pyrrhotite", "magnetite", "hematite"))
aFe <- affinity("Fe+2" = c(0, 12), O2 = c(-40, -16), T = 400, P = 2000)
dFe <- diagram(aFe, xlab = xlab)
title(paste("Only Fe; 1° balance:", dFe$balance))
label.plot("A")

# Only Cu-bearing minerals
species(c("covellite", "chalcocite", "tenorite", "cuprite"))
aCu <- affinity(aFe)  # argument recall
dCu <- diagram(aCu, xlab = xlab)
title(paste("Only Cu; 1° balance", dCu$balance))
label.plot("B")

# Only Fe- AND Cu-bearing minerals
species(c("chalcopyrite", "bornite"))
aFeCu <- affinity(aFe)
dFeCu <- diagram(aFeCu, xlab = xlab, balance = "H+")
title(paste("Only Fe+Cu; 1° balance:", dFeCu$balance))
label.plot("C")

### SECONDARY balancing

# Fe- or Cu-bearing minerals
ad1 <- duplex(dFe, dCu, balance = "H+")
d1 <- diagram(ad1, xlab = xlab, balance = 1)
title("Only Fe or Cu; 2° balance: H+")
label.plot("D")

# All minerals
d1$values <- c(dFe$values, dCu$values)
ad2 <- duplex(d1, dFeCu, balance = "H+")
abbrv <- info(ad2$species$ispecies)$abbrv
diagram(ad2, xlab = xlab, balance = 1, names = abbrv)
title("All Fe or Cu; 2° balance: H+")
label.plot("E")

db <- describe.basis(ibasis = 3)
leg <- lex(lTP(400, 2000), db)
legend("bottomleft", legend = leg, bty = "n")
```
</div>

We first define basis species to contain both Cu- and Fe-bearing species.
The x-axis is the ratio of activities of Fe^+2^ and Cu^+^; the label is made with `ratlab()`.
```{r duplex, eval = FALSE, echo = 5:6}
```

We then calculate the diagrams for the primary balancing coefficients, for the groups of only Fe-, only Cu-, and only Fe+Cu-bearing minerals.
It is obvious that the first two systems are balanced on Fe and Cu, respectively, but the third has a somewhat unusual balance: H^+^.
See Reaction 4 of @MH85 for an example.
```{r duplex, eval = FALSE, echo = 10:30}
```

Now comes the secondary balancing, where all reactions, not only that between bornite and chalcopyrite, are balanced on H^+^.
We first combine the affinities for the Fe- or Cu-bearing minerals to make diagram D.
Note that after secondary balancing with `duplex()`, the argument `balance = 1` should be used in `diagram()` to prevent further balancing.
This is because `duplex()` preserves the primary balancing for Fe- and Cu-bearing minerals (internally the "plotvals" components of `dFe` and `dCu`).

Then we combine diagrams D and C to make the final diagram in E.
The fields in this diagram are labeled with mineral abbreviations from the OBIGT database.
```{r duplex, eval = FALSE, echo = 33:43}
```

```{r duplex, echo = FALSE, results = "hide", message = FALSE, fig.width = 6.5, fig.height = 5, out.width = "100%", fig.align = "center", pngquant = pngquant}
```

*Challenge*: Although the diagram here is drawn only for H~2~S in the basis species, take it a step further and make a mosaic diagram to account for the stability of HSO~4~^-^ at high oxygen fugacity.

## Other Possibilities

Conceptually, the methods described above treat different metal-bearing elements as parts of distinct chemical systems that are then joined together.
Other methods may be more suitable for considering multiple metals (or other elements) in one system.

### Balancing on a Non-Metal

As shown in the **secondary balancing** example, there is no requirement that the balancing coefficients come from a metal-bearing species.
It is possible to make diagrams for minerals with different metallic elements simply by using a non-metallic element as the primary balance.
Here is an example for the Cu-Fe-S-O-H system based on Figure 3 of @MH85.
The reactions are balanced on O~2~, which means that no O~2~ appears in the reaction between any two minerals, but Fe^+2^ and/or Cu^+^ can be present, depending on the chemical composition.
Saturation limits are shown for species that have no O~2~ in their formation reactions.

```{r non-metal, results = "hide", message = FALSE, fig.width = 6, fig.height = 5, out.width = "80%", fig.align = "center", pngquant = pngquant}
basis(c("Fe+2", "Cu+", "hydrogen sulfide", "oxygen", "H2O", "H+"))
basis("H2S", 2)
species(c("pyrite", "magnetite", "hematite", "covellite", "tenorite",
          "chalcopyrite", "bornite"))
a <- affinity("Cu+" = c(-8, 2, 500), "Fe+2" = c(-4, 12, 500), T = 400, P = 2000)
d <- diagram(a, xlab = ratlab("Cu+"), ylab = ratlab("Fe+2"), balance = "O2")
title(paste("Cu-Fe-S-O-H; 1° balance:", d$balance))
# Add saturation lines
species(c("pyrrhotite", "ferrous-oxide", "chalcocite", "cuprite"))
asat <- affinity(a)  # argument recall
diagram(asat, type = "saturation", add = TRUE, lty = 2, col = 4)
legend("topleft", legend = lTP(400, 2000), bty = "n")
```

In some ways this is like the inverse of the **mosaic series** example.
There, reactions were balanced on Fe or Cu, and *f*~O<sub>2</sub>~ and pH were used as plotting variables.
Here, the reactions are balanced on O~2~ and implicitly on H^+^ through the activity ratios with *a*~Fe^+2^~ and *a*~Cu^+^~, which are the plotting variables.

More common diagrams of this type are balanced on Si or Al.
See `demo(saturation)` for an example in the H~2~O-CO~2~-CaO-MgO-SiO~2~ system.

### Mosaic Combo

Instead of adding minerals with different metals by stacking mosaic diagrams (**mosaic series**), it may be possible to include two different metals in the basis species and formed species.
The `mosaic()` and `equilibrate()` functions can be combined to balance on two different elements.
The example here is for N and C rather than metals.
Thanks to Kirt Robinson for the feature request and test case.

First, the `mosaic()` command calculates equilibrium activities of NH~3~ and NH~4~^+^ for a given total activity of N in the basis species, and calculates the corresponding affinities of the formed species.
Then, the `equilibrate()` command calculates equilibrium activities of the formed species for given total activity of C, and combines them with the activities of the changing basis species (NH~3~ and NH~4~^+^).

```{r mosaic-combo, results = "hide", message = FALSE, fig.width = 6, fig.height = 5, out.width = "80%", fig.align = "center", pngquant = pngquant}
loga_N <- -4
loga_C <- -3
basis(c("CO2", "NH3", "O2", "H2O", "H+"))
basis("NH3", loga_N)
species(c("acetamide", "acetic acid", "acetate"))
m <- mosaic(c("NH3", "NH4+"), pH = c(0, 14))
e <- equilibrate(m, loga.balance = loga_C)
diagram(e, ylim = c(-10, -2))
tN <- paste("log(total N in basis species) =", loga_N)
tC <- paste("log(total C in formed species) =", loga_C)
title(main = paste(tN, tC, sep = "\n"), font.main = 1)
```

The diagram shows the expected ionization of acetic acid and NH~3~ at different pHs.
The appearance of acetamide (CH~3~CONH~2~) is a consequence of the interaction between the N-bearing and C-bearing species, and is analogous to the formation of a multi-metal complex.


## Document History

* 2020-07-15 First version.

## References
